---
title: "Isolating data with dplyr"
author: "Rachel Stuckwisch"
date-modified: last-modified
format:
  html:
    code-folding: show
    df-print: paged
    fig-caption: true
    fig-height: 4
    fig-width: 7
    highlight: tango
    theme: cosmo
    toc: true
    toc-float: true
---

# Overview

NHANES (The National Health and Nutrition Examination Survey) was designed to assess the health and nutritional status of the US population and is conducted by the National Center for Health Statistics of the Centers for Disease Control and Prevention. Since 1999-2000, NHANES has been conducted in two-year cycles. For each cycle, potential participants are identified through stratified, multistage probability sampling of the non-institutionalized US population. In this set of exercises, we will use the ten cycles conducted from 1999-2000 through 2017-2018. 

## Import

```{r}
#| echo: false

# Load packages 
library(tidyverse)  # includes dplyr - filter / select
library(gt)

nhanes <- read_rds('data/NHANES_analysis_pre_exclusions.rds')

# trying to print all rows will make df_print: paged cause a pandoc error.
# don't modify the line below! It may break knitr.
nhanes[1:10, ]
  
```


## Data dictionary

Review this briefly and use it as a reference to engage with the exercises below.

```{r}
#| echo: false

# A data dictionary is more comprehensive than these two columns,
# but these two columns are the core components of a data dictionary.
# I highly recommend you only engage with data when there is a 
# data dictionary available. If a collaborator cannot share one of
# these with you when they ask you to analyze your data, then they 
# should not be your collaborator.

nhanes_descr <- c(
  "seqn"              = "SP identifier",
  "exam"              = "NHANES exam year",
  "psu"               = "primary sampling unit",
  "strata"            = "survey strata",
  "wts_mec_2yr"       = "survey weights",
  "exam_status"       = "How did SP engage with exam?",
  "age"               = "SP age, years",
  "age_group"         = "SP age group, years",
  "sex"               = "SP sex",
  "race_ethnicity"    = "SP race and/or ethnicity",
  "education"         = "SP education",
  "income_hh"         = "SP household income",
  "pregnant"          = "was SP pregnant at time of exam?",
  "bp_sys_mmhg"       = "SP systolic blood pressure, mm Hg",
  "bp_dia_mmhg"       = "SP diastolic blood pressure, mm Hg",
  "n_msr_sbp"         = "Number of valid systolic BP readings",
  "n_msr_dbp"         = "Number of valid diastolic BP readings",
  "bp_controlled"     = "Did SP have controlled BP? (<140/90 mm Hg)",
  "acr_mgg"           = "SP albumin-to-creatinine ratio, mg/g",
  "albuminuria"       = "Did SP have albuminuria? (ACR > 30 mg/g)",
  "chol_hdl_mgdl"     = "SP HDL-cholesterol, mg/dl",
  "chol_total_mgdl"   = "SP total cholesterol, mg/dl",
  "health_insurance"  = "SP health insurance status",
  "bp_high_aware"     = "SP ever told by Dr: 'you have high blood pressure'?",
  "bp_meds"           = "SP currently using antihypertensive medication?",
  "hc_usual_facility" = "SP has a usual healthcare facility?",
  "hc_visit_1yr"      = "SP visited their healthcare facility last year?"
)

# abbreviations need to be placed in the footnote of the table
abbrevs <- c("SP = survey participant", 
  "BP = blood pressure", 
  "HDL = high density lipoprotein") |> 
  paste(collapse = '; ')

# the enframe function transforms a vector into a tibble,
enframe(nhanes_descr) |> 
  gt(rowname_col = "name") |>
  tab_stubhead(label = 'Variable name') |> 
  cols_label(value = 'Variable description') |> 
  cols_align('right') |> 
  tab_source_note(source_note = abbrevs) |> 
  tab_header(title = 'Description of NHANES data')

# the paste() function is used here. To give a brief summary,
# paste() combines string vectors. In this case, we are using 
# paste to collapse a vector and separate each component by '; '

```


# Pre-requisites

Before starting these exercises, you should have a good understanding of 

1. The [Transform Tables](https://posit.cloud/learn/recipes){.external target="_blank"} Recipes, specifically filtering (6th) and selecting (1st)

1. Chapter 3 of [R for Data Science](https://r4ds.hadley.nz/data-transform){.external target="_blank"}

# Exercise 1

Suppose we are conducting a study and aim to include NHANES participants meeting the following conditions: 

1. Age 18+
1. Completed the NHANES interview and examination
1. Not pregnant at time of exam
1. 1+ systolic and diastolic BP measurement
1. Complete information on BP medication use.

Create a `tibble` with two columns: 

- `inclusion`: a numeric vector taking values 1, 2, 3, 4, and 5

- `description`: a character vector comprising the inclusion criteria above.

```{r}

prob1 <- tibble("inclusion" = 1:5, 
                "description" = c("Age 18+",
                                  "Completed the NHANES interview and examination",
                                  "Not pregnant at time of exam",
                                  "1+ systolic and diastolic BP measurement",
                                  "Complete information on BP medication use"))

prob1
```

# Exercise 2

Apply the `filter` function to apply each inclusion criteria, in the same sequence as I listed earlier, to the NHANES data. Because we need to track how many participants were included after each step, I would recommend creating five separate datasets, like so: 

```{r}

# your code will fill in the ... with conditions that 
# align with the inclusion criteria given above.
e1 <- filter(nhanes, age >= 18)
e2 <- filter(e1, exam_status == "interview and exam")
e3 <- filter(e2, pregnant == "No" | is.na(pregnant))
e4 <- filter(e3, n_msr_sbp != 0 & !is.na(n_msr_sbp) & n_msr_dbp != 0 & !is.na(n_msr_dbp))
e5 <- filter(e4, !is.na(bp_meds))

# hints: 
# 
# use the data dictionary to identify relevant variables.
#
# is.na() returns TRUE when a value is missing. Run this example to verify
# x <- c(1, NA, 3)
# is.na(x)
#
# for e3, be careful. Pregnant is NA for some female participants.
# In this case, we only want to apply the exclusion if we are certain
# that a person was pregnant, so what we really need is to include 
# participants who have pregnant == 'No' OR pregnant is missing.
# think about how you could use the is.na() function.
# 
# for e4 and e5, think about how you could use the is.na() function
# in combination with the NOT (!) operator. For example, !condition
# returns TRUE is the condition is NOT TRUE. 

```

Create a numeric vector of length 5 that contains the number of rows in each dataset you created (i.e., `nrow(e1)`, ..., `nrow(e5)`). Attach that vector to the `tibble` you created in exercise 1, and name it `n`. 

```{r}
# make a 5x1 vector
ndata <- c(nrow(e1), nrow(e2), nrow(e3), nrow(e4), nrow(e5)) 
# turn vector into a tibble
ndata <- as_tibble(ndata)
# rename tibble column (value -> n)
  # https://dplyr.tidyverse.org/reference/rename.html
  # rename(data, new = old) 
ndata <- rename(ndata, n = value )

prob2 <- bind_cols(prob1, ndata)

prob2

# Hints. 
# Remember that you can access or add columns to lists (and transitively, 
  # data frames) using $. for example, nhanes$new_col <- 1 adds a new column 
  # of 1's to nhanes.


```

# Exercise 3

Apply `gt()` to form a publication-ready table describing your inclusions.

```{r}

tbl3 <- prob2 |>
  select(description, n) |>
  gt(rowname_col = "description") |>
  tab_stubhead(label = "Inclusion criteria") |>
  cols_label(n = "No. of participants") |>
  tab_footnote(
    footnote = "The inclusion criteria were applied sequentially",
    locations = cells_stubhead()) |>
  tab_source_note("BP = blood pressure; No. = number") |>
  cols_align(align = "center",
             columns = n)
tbl3

# Hint: My solution looks like this:
#
# nhanes_sample |>            [start with the solution from problem 2]
#   select() |>               [remove one column]
#   gt(rowname_col = ???) |>  [initiate the table]
#   tab_stubhead() |>         [label table stub]
#   cols_label() |>           [label columns]
#   ...                        [do other things]

saveRDS(tbl3, "solutions/03_solution.rds")

```

# Exercise 4

Using your final sample of 51,761 NHANES participants, identify whether the highest value of systolic BP was provided by a male or female in each exam year. To do this, you should apply a grouped filter (i.e., use `group_by()` and then `filter()`).

- The groups should be defined by `exam`.

- The filter step should keep rows that are equal to the maximum systolic BP value.

- A final step should apply the select function to keep only the `exam`, `bp_sys_mmhg`, and `sex` columns. 

```{r}

#maximum systolic BP values
prob4_max <- group_by(e5, exam) |>
  filter(bp_sys_mmhg == max(bp_sys_mmhg)) |>
  select(exam, bp_sys_mmhg, sex)
prob4_max

# minimum systolic BP values
prob4_min <- group_by(e5, exam) |>
  filter(bp_sys_mmhg == min(bp_sys_mmhg)) |>
  select(exam, bp_sys_mmhg, sex)
prob4_min

```

Do you notice anything peculiar? Try looking at the minimum systolic BP values instead. Do you notice any peculiar patterns here?


