---
title: "Table Exercise"
author: "YOUR NAME"
date: "Today's date"
format:
  html:
    code_folding: show
    df-print: paged
    fig-caption: true
    fig-height: 4
    fig-width: 7
    highlight: tango
    theme: cosmo
    toc: true
    toc-float: true
---

## Problem 1

- Get some practice working with `gt`

- Use what you learned in the intro to markdown file to create the table below:

**Hints**: 

- Note, we're using rows 1, 2, 3, 51, 52, 53, 101, 102, and 103 of the `iris` dataset

- The `iris` dataset is already loaded in your R environment, no need to import it


```{r}

library(gt)
readr::read_rds('solutions/iris_table.rds')

```

### Challenge yourself

There are a couple of things not discussed in the vignette that are applied here, not necessary but if you're feeling ambitious...

1.   Proper case on the species names\
    __Hint__ - Not a gt function, do this early
2.   Center aligning the columns\
    __Hint__ - Use the `gt::col_align()` function
3.   Using Georgia as the table font\
    __Hint__ - There are many `opt_` functions in `gt`, maybe there's one for table fonts?
4.   Removing bootstrap striping
    
:::{.callout-caution} 
#4 will take some digging in Quarto, good luck!
:::

```{r}

#INSERT YOUR CODE HERE - DON'T FORGET TO COMMENT



```


