---
title: "Joining data"
author: "Rachel Stuckwisch"
date-modified: last-modified
format:
  html:
    code_folding: show
    df-print: paged
    fig-caption: true
    fig-height: 4
    fig-width: 7
    highlight: tango
    theme: cosmo
    toc: true
    toc-float: true
---

# Overview

These exercises will help you practice applying functions in the `join` family and the `stringr` and `glue` packages. You will use two datasets that describe a TV show called 'The Office'.

## Setup

Packages used for this exercise set are loaded below.

```{r}
#| echo: false
#| warning: false
#| error: false

library(tidyverse)
library(glue)
library(gt)

```

## Data dictionary

The analysis will involve two datasets:

-   `office_ratings.csv`: the IMDB rating for each episode.

-   `office_director_and_writers.csv` the writers and director of each episode.

```{r}

ratings_descr <-
  c("season"      = "Season that the episode aired",
    "episode"     = "Episode number, relative to Season start",
    "title"       = "Title of the episode",
    "imdb_rating" = "Mean IMDB rating",
    "total_votes" = "Total number of IMDB members who rated the episode",
    "air_date"    = "Date that the episode aired")

drctr_wrtr_descr <- 
  c("season"   = "Season that the episode aired", 
    "title"    = "Title of the episode", 
    "director" = "Name of the episode's director", 
    "writer"   = "Name of contributing writer")

data_names <- c("ratings", "director and writers")

tbls <- list(ratings_descr, drctr_wrtr_descr) |> 
  map2(.y = data_names,
    ~ enframe(.x) |>
      gt(rowname_col = "name") |>
      tab_stubhead(label = 'Variable name') |> 
      cols_label(value = 'Variable description') |> 
      cols_align('left') %>% 
      tab_header(title = glue('Description of {.y} data'))
  )

tbls[[1]]

```

```{r}

tbls[[2]]

```

## Import

You can import the two datasets using the `read_csv()` function. Note that your project directory is the root of the file path.

```{r}

# Read in both the 'office_ratings.csv' file and 
# the 'office_director_and_writers.csv' file.

```

# Problem 1

Conduct exploratory data analysis of the `office_ratings` data and produce the figure shown below.


```{r}

# Your code goes here

```


```{r, out.width='100%'}

readr::read_rds('solutions/01_solution.rds')

```

# Problem 2

Identify the episode names that are present in `office_ratings` but not in `office_director_and_writers`. Then, identify the episode names that are present in `office_director_and_writers` but not in `office_ratings`. I saved each vector of character values in the list printed below.


```{r}

# Your code goes here

```


```{r}

readr::read_rds('solutions/02_solution.rds')

```

# Problem 3

Design and implement a strategy to harmonize the episode names between these two data sets so that they can be joined. When you are finished, `anti_join()`ing the two data sets should produce an empty data set.

Once the names are harmonized, left join the `office_director_and_writers` data with the `office_ratings` data

Hints: 

- I used the following functions to harmonize episode names: `tolower()`, `str_remove_all()`, and `str_replace()`.

- Don't worry if it takes more than a few steps to get everything harmonized. Just take it one step at a time.


```{r}

# Your code goes here

```


```{r}

readr::read_rds('solutions/03_solution.rds')

```

# Problem 4

Summarize the merged data by computing 

1. the number of episodes that each writer contributed to.
1. the mean rating of episodes by writer.

Using the summarized data, create a new column that indicates who is a `top_writer`. The criteria for being identified as a `top_writer` are having a mean episode rating > 8.6 or writing > 20 episodes.


```{r}

# Your code goes here

```


```{r}

readr::read_rds('solutions/04_solution.rds')

```

# Problem 5

Merge the data you created in problem 4 with the data you previously merged in problem 3, and then create a dataset that includes the top 5 episodes for each `top_writer`, arranged by rating. If the `top_writer` wrote k < 5 episodes, just show the k episodes that the writer contributed to. 

Notes: 

- `group_by()`, `arrange()`, and `slice()` work extremely well together.


```{r}

# Your code goes here

```


```{r}

readr::read_rds('solutions/05_solution.rds')

```










