---
title: "Data visualization tools"
author: "Rachel Stuckwisch"
date-modified: last-modified
format:
  html:
    code-folding: show
    df-print: paged
    fig-caption: true
    fig-height: 6
    fig-width: 10
    highlight: tango
    theme: cosmo
    toc: true
    toc-float: true
---

```{r}
#| echo: false

```

# Overview

The `supermarket` data are synthetic but realistic observations of customer interactions in various counties. The exercises in this document are based on a [tutorial](https://uc-r.github.io/cleveland-dot-plots) written on the UC Business Analytics blog

## Import

```{r}
#| echo: false
# Load packages 
library(tidyverse)  # includes dplyr - filter / select
library(gt)
library(readxl)
library(janitor)


supermarket <- read_excel(
  # the file location
  path = "data/Supermarket Transactions.xlsx",
  # the sheet name to pull data from
  # you can also use numeric order of sheets
  sheet = "Data"
) 

  
```

## Data dictionary

Review this briefly and use it as a reference to engage with the exercises below.

```{r}

supermarket_descr <-
  c(
    "Transaction"        = "transaction number",
    "Purchase Date"      = "date of purchase",
    "Customer ID"        = "customer identification",
    "Gender"             = "--",
    "Marital Status"     = "--",
    "Homeowner"          = "--",
    "Children"           = "number of kids at home",
    "Annual Income"      = "income of customer",
    "City"               = "--",
    "State or Province"  = "--",
    "Country"            = "Canada, Mexico, or USA",
    "Product Family"     = "drink, food, or non-consumable",
    "Product Department" = "the item's department",
    "Product Category"   = "the item's type",
    "Units Sold"         = "number of items sold",
    "Revenue"            = "money generated by the transaction"
  )

# the enframe function transforms a vector into a tibble,
enframe(supermarket_descr) |> 
  gt(rowname_col = "name") |>
  tab_stubhead(label = 'Variable name') |> 
  cols_label(value = 'Variable description') |>
  cols_align('right') |>
  tab_header(title = 'Description of Super Market data')

```

# Exercise 1

Use the `janitor` package to clean your dataset's names. Then, 

- group your data by city and gender,

- compute the sum of revenue for each group

In the summarized data, 

- (hint: `fct_recode`) convert gender to a factor with categories

    + 'Male customers'
    
    + 'Female customers'
    
- (hint: `fct_reorder`) convert `city` to a factor with levels ordered by revenue.

- Don't forget to `ungroup` your data after summarizing!

```{r}

# your code goes here

```

Your solution should look like this:

```{r}
#| echo: false

read_rds('solutions/01_solution.rds')

```

# Exercise 2

Using the summarized data from exercise 1, re-create the figure presented during lecture:

```{r}

# Your code goes here

```

Your solution should look like this:

```{r}
#| echo: false

read_rds('solutions/02_solution.rds')

```

# Exercise 3

create two datasets: `left_label` and `right_label`:

- `left_label` should contain the rows of your summarized data from exercise 1 that are on the left hand side of the plot from exercise 2. 

- `right_label` should contain the rows of your summarized data from exercise 1 that are on the right hand side of the plot from exercise 2. 

```{r}

# Your code goes here

```


Your solution should look like this:

```{r}
#| echo: false

read_rds('solutions/03_solution.rds')

```

# Exercise 4

Add two text geometries to your figure from exercise 1. Use `nudge_x` to shift your text to the right and left of the points. 

```{r}

# Your code goes here

```

Your solution should look like this:

```{r}
#| echo: false

read_rds('solutions/04_solution.rds')

```

# Exercise 5

- Create a dataset called `big_diff` that shows the absolute value of the difference in revenue generated between male customers and female customers in each city.

- Filter this dataset so that only cities with difference greater than \$750 are retained.

- Filter the `right_label` data from exercise 3 to retain only the rows where `city` is in `big_diff$city` values.

- Create a dataset called `highlight` that filters your summarized data from exercise 1 to retain only the rows where `city` is in `dig_diff$city`

```{r}


# Your code goes here

```

Your filtered `right_label` and `highlight` datasets should look like this:

```{r}
#| echo: false

read_rds('solutions/05_solution.rds')

```

# Exercise 6

Create a plot similar to that from exercise 4, but this time 

- make lines and points have `alpha` of 1/2. 

- color the lines `'grey'`

- use only the `right_label` data to add text to the plot

- Set `label` to be the difference in revenue rather than revenue. 

```{r}

# Your code goes here

```

Your solution should look like this:

```{r}
#| echo: false

read_rds('solutions/06_solution.rds')

```

# Exercise 7

- Add lines and points to the figure you created in exercise 6, using the `highlight` data you created in exercise 5.

- Use `scale_x_continuous`, setting `labels` to `scales::dollar` so that your x-axis will show dollar signs.

- Modify your x-axis and y-axis labels. 

```{r}

# Your code goes here

```

Your solution should look like this:

```{r}
#| echo: false

read_rds('solutions/07_solution.rds')

```

# Exercise 8

Apply color palettes, theme elements, and legend manipulations to make your figure clear and concise. 

```{r}

# Your code goes here

```

Your solution should look like this:

```{r}
#| echo: false

read_rds('solutions/08_solution.rds')

```



