---
title: "Reshaping Data"
author: "Rachel Stuckwisch"
date-modified: last-modified
format:
  html:
    code_folding: show
    df-print: paged
    fig-caption: true
    fig-height: 4
    fig-width: 7
    highlight: tango
    theme: cosmo
    toc: true
    toc-float: true
    knitr: 
      opts_chunk: 
        echo: true
---

# Overview

These exercises will help you practice applying `pivot_wider` and `pivot_longer`, and will also require some use of skills you have previously learned. No new data (other than toy examples) will be introduced.

## Setup

I am not loading packages this time. You can load your own as you determine what tools are needed


```{r}
#| echo: false

# Load packages 

# I am not loading packages this time.
# You can load your own as you determine what tools are needed

```

# Exercise 1

Convert the given dataset from wide to long form

```{r}

preg <- tibble::tribble(
  ~pregnant, ~male, ~female,
  "yes",     NA,    10,
  "no",      20,    12
)

```

Your solution should look like this:

```{r}
#| echo: false

readr::read_rds('solutions/01_solution.rds')

```

# Exercise 2

Convert the long dataset you created in exercise 1 into a wide dataset with 

- `sex` a column indicating male or female sex.

- `preg_yes` count of pregnancies for the given sex

- `preg_no` count of non-pregnancies for the given sex.

```{r}

# Your code goes here

```

Your solution should look like this:

```{r}
#| echo: false

readr::read_rds('solutions/02_solution.rds')

```

# Exercise 3

Convert the following data into long form, with one column for `sbp`, `dbp`, and `hr` (heart rate). 

```{r}
bp <- tibble::tribble(
  ~subjid, ~sbp_1, ~sbp_2, ~dbp_1, ~dbp_2, ~hr_1, ~hr_2,
  'a',      123,    NA,    68,     72,     NA,    70,    
  'b',      140,    139,    NA,     82,     NA,    65
)
```


```{r}

# Your code goes here

```

Your solution should look like this:

```{r}
#| echo: false

readr::read_rds('solutions/03_solution.rds')

```

# Exercise 4

Using the data you created in exercise 3, add three new columns that measure the number of valid `sbp`, `dbp`, and `hr` measurements for each `subjid`.

```{r}

# Your code goes here

```

Your solution should look like this:

```{r}
#| echo: false

readr::read_rds('solutions/04_solution.rds')

```

# Exercise 5

- Read in the `nhanes_long.rds` file, located in the `data` directory.

- Pivot the data into a wider format and create the table shown below.

```{r}

# Your code goes here

```


```{r}
#| echo: false

readr::read_rds('solutions/05_solution.rds')

```


